\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2021-11-28 main.Rmd
introduction revised

**Spatio-temporal data + examples**  
Spatio-temporal data record changes of variables in spatially separated regions across time. In this article, we consider spatio-temporal vector data, which are recorded in a fixed interval. Examples of this type of data include the house price of a city or county, climate measures from weather stations in a country, and river level data from electronic gauges \textcolor{red}{more examples here}. \newline

<!-- and are point based, characterised by longitude and latitude, in the spatial aspect -->

**point to common feature: spatial level variables + time level variables relates to Table 13 in Tidy data paper**  
Usually, this type of spatio-temporal data does not necessarily come in as a single table. Tidy data principle [@tidydata] prescribes each type of observational unit to form a table. This would suggest two tables to store the data, one for spatial-level and one for temporal-level. The Table 13 in @tidydata presents a structure like this and the author argues the lack of tools to work with relational data like such. \newline

**My proposal**  
Recent software development in R has proposed several relational data structure: `tidygraph`[@tidygraph] for graph manipulation, `dm` [@dm] for relational data model, while spatio-temporal data could benefit from having its own relational data structure that identifies the essential spati0-temporal element: site identifier, time identifier, geographic coordinates. In this paper, we propose a tidy data structure for vector spaio-temporal data.  \newline

**Section division**  
The rest of the paper will be divided as follows: Section 2 reviews the existing data structure for spatio, temporal, and spatio-temporal data. Section 3 presents a new data structure for spatio-temporal data: cubble. Then the paper introduces the workflow of data manipulation and visualisation with the cubble structure in Section 4. Section 5 gives some examples on how common spatial and temporal manipulations are performed with cubble and how static and interactive visualisation help to understand climate and [...] data. \newline

# Existing data structure for spatio and temporal data {#lit-review}

**Existing packages**  
Many data structures have been proposed for spatial and temporal data, but not many for spatio-temporal data. One of the reason could be the inherent different levels of information make it inefficient to store in the same table. `spacetime`[@spacetime] proposed four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT) based on underlying spatial structure `sp` [@sp] and temporal structure `xts`[@xts]. `spatstat` [@spatstat] implements a `ppp` class for point pattern data. More recent package `stars` [@stars] uses a spatiotemporal array to store the data and the array structure has its influence from `cubelyr` [@cubelyr], a dplyr data cube backend. \newline

**Criteria of a data structure we want**

With recent development in the R community, `sf` [@sf] and `tsibble`[@tsibble] have replaced `sp` and `xts` to be the convention structure for spatial and temporal data. One reason for their popularity is its integration with tidyverse ecosystem, making them intuitive and easy to adopt. For spatio-temporal data, we hope to build a data structure has the following features:

  1) A data structure that handles spatial and temporal dimension in a relational structure. This derives from the 3rd tidy data principal.
  2) An intuitive and easy to use interface that fits into the tidyverse ecosystem, and
  3) Compatibility with the latest spatial and temporal data structure. This would give users the flexibility to use the work in existing packages.

>>>>

**Motivation**

Many data structures have been proposed for spatial (`sf` by @sf) and temporal (`tsibble` by @tsibble) data in the R community, while less has been done for spatio-temporal data. The lack of such tools could potentially  because analysts usually treat the spatial and temporal dimension pf the data separately, without realising the need to create a new data structure. While this approach follows the third tidy data principal [@tidydata] (*Each type of observational unit forms a table*), analysts always need to manually join results from different observational units or combining multiple tables into one for downstream analysis. This additional step doesn't add new operations into the data but can be error prone.  \newline

**Existing packages**

Currently, available spatio-temporal data structure in R includes: `spacetime`[@spacetime], which proposes four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT). The data structure it uses is based on `sp` [@sp] and `xts`[@xts], both of which has been replaced by more recent implementations. `spatstat` [@spatstat] implements a `ppp` class for point pattern data; and more recent, `stars` [@stars] implements a spatio-temporal array with the dplyr's data cube structure `cubelyr` [@cubelyr] as its backend. While these implementations either store spatial and temporal variables all in a single table, hence duplicate the spatial variables for each temporal unit; or split them into two separate tables that has the problem of manually joining, mentioned in the previously. None of these packages enjoy both the benefits of being able to separate manipulation in the two dimensions while also keep the data object as a whole. This create a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: the Ireland wind data from `gstat` is an classic example data that splits variables into spatial (`wind.loc`) and temporal (`wind`) dimension;Bureau of Meteorology (BoM) provides climate observations that are widely applied in agriculture and ecology study; air pollution data. \newline

**Our new data structure for spatio-temporal data**  
This paper describes the implementation of a new spatio-temporal data structure: `cubble`. `cubble` implements a relational data structure that uses two forms to manage the switch between spatial and temporal dimension. With this structure, users can manipulate the spatial or temporal dimension separately, while leaves the linking of two dimensions to `cubble`. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link]. \newline

**Section division**

The rest of the paper will be divided as follows: [complete when the paper structure is more solid]

<!-- Section 2 reviews the existing data structure for spatio, temporal, and spatio-temporal data. Section 3 presents a new data structure for spatio-temporal data: cubble. Then the paper introduces the workflow of data manipulation and visualisation with the cubble structure in Section 4. Section 5 gives some examples on how common spatial and temporal manipulations are performed with cubble and how static and interactive visualisation help to understand climate and [...] data. \newline -->

### 2021-11-28 main.tex
introduction revised

In spatio-temporal data analysis, there's a need for a new
spato-temporal data structure. The requirement for such a tool is
necessary given the ubiquity of spatio-temporal data in the wild.
Example of this includes house price of a city or county, climate
station data from the government,

@@

Many data structures have been proposed for spatial and temporal data,
but not many for spatio-temporal data. One of the reason could be the
inherent different levels of information make it inefficient to store in
the same table. \texttt{spacetime}\citep{spacetime} proposed four
space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI),
and trajectory (STT) based on underlying spatial structure \texttt{sp}
\citep{sp} and temporal structure \texttt{xts}\citep{xts}.
\texttt{spatstat} \citep{spatstat} implements a \texttt{ppp} class for
point pattern data. More recent package \texttt{stars} \citep{stars}
uses a spatiotemporal array to store the data and the array structure
has its influence from \texttt{cubelyr} \citep{cubelyr}, a dplyr data
cube backend. \newline

\textbf{Our new data structure for spatio-temporal data}

In this paper, a new spatio-temporal data structure, \texttt{cubble}, is
introduced. \texttt{cubble} implements a relational data structure that
splits the data into its spatial and temporal component. With this
structure, users can separately manipulate the spaital or temporal
component of the data while keep the two dimension linked. The software
is available from the Comprehensive R Archive Network (CRAN) at {[}CRAN
link{]}. \newline

@@

The rest of the paper will be divided as follows: Section 2 reviews the
existing data structure for spatio, temporal, and spatio-temporal data.
Section 3 presents a new data structure for spatio-temporal data:
cubble. Then the paper introduces the workflow of data manipulation and
visualisation with the cubble structure in Section 4. Section 5 gives
some examples on how common spatial and temporal manipulations are
performed with cubble and how static and interactive visualisation help
to understand climate and {[}\ldots{]} data. \newline

\textbf{Criteria of a data structure we want}

With recent development in the R community, \texttt{sf} \citep{sf} and
\texttt{tsibble}\citep{tsibble} have replaced \texttt{sp} and
\texttt{xts} to be the convention structure for spatial and temporal
data. One reason for their popularity is its integration with tidyverse
ecosystem, making them intuitive and easy to adopt. For spatio-temporal
data, we hope to build a data structure has the following features:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  A data structure that handles spatial and temporal dimension in a
  relational structure. This derives from the 3rd tidy data principal.
\item
  An intuitive and easy to use interface that fits into the tidyverse
  ecosystem, and
\item
  Compatibility with the latest spatial and temporal data structure.
  This would give users the flexibility to use the work in existing
  packages.
\end{enumerate}

\textbf{Spatio-temporal data + examples}\\
Spatio-temporal data record changes of variables in spatially separated
regions across time. In this article, we consider spatio-temporal vector
data, which are recorded in a fixed interval. Examples of this type of
data include the house price of a city or county, climate measures from
weather stations in a country, and river level data from electronic
gauges \textcolor{red}{more examples here}. \newline

\textbf{point to common feature: spatial level variables + time level
variables relates to Table 13 in Tidy data paper}\\
Usually, this type of spatio-temporal data does not necessarily come in
as a single table. Tidy data principle \citep{tidydata} prescribes each
type of observational unit to form a table. This would suggest two
tables to store the data, one for spatial-level and one for
temporal-level. The Table 13 in \citet{tidydata} presents a structure
like this and the author argues the lack of tools to work with
relational data like such. \newline

\textbf{My proposal}\\
Recent software development in R has proposed several relational data
structure: \texttt{tidygraph}\citep{tidygraph} for graph manipulation,
\texttt{dm} \citep{dm} for relational data model, while spatio-temporal
data could benefit from having its own relational data structure that
identifies the essential spati0-temporal element: site identifier, time
identifier, geographic coordinates. \newline

\hypertarget{lit-review}{%
\section{Existing data structure for spatio and temporal
data}\label{lit-review}}

>>>>

Many data structures have been proposed for spatial (\texttt{sf} by
\citet{sf}) and temporal (\texttt{tsibble} by \citet{tsibble}) data in
the R community, while less has been done for spatio-temporal data. The
lack of such tools could potentially because analysts usually treat the
spatial and temporal dimension pf the data separately, without realising
the need to create a new data structure. While this approach follows the
third tidy data principal \citep{tidydata} (\emph{Each type of
observational unit forms a table}), analysts always need to manually
join results from different observational units or combining multiple
tables into one for downstream analysis. This additional step doesn't
add new operations into the data but can be error prone. \newline

@@

Currently, available spatio-temporal data structure in R includes:
\texttt{spacetime}\citep{spacetime}, which proposes four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). The data structure it uses is based on \texttt{sp}
\citep{sp} and \texttt{xts}\citep{xts}, both of which has been replaced
by more recent implementations. \texttt{spatstat} \citep{spatstat}
implements a \texttt{ppp} class for point pattern data; and more recent,
\texttt{stars} \citep{stars} implements a spatio-temporal array with the
dplyr's data cube structure \texttt{cubelyr} \citep{cubelyr} as its
backend. While these implementations either store spatial and temporal
variables all in a single table, hence duplicate the spatial variables
for each temporal unit; or split them into two separate tables that has
the problem of manually joining, mentioned in the previously. None of
these packages enjoy both the benefits of being able to separate
manipulation in the two dimensions while also keep the data object as a
whole. This create a gap in the software development. The requirement
for such a tool is important given the ubiquity of spatio-temporal
vector data in the wild: the Ireland wind data from \texttt{gstat} is an
classic example data that splits variables into spatial
(\texttt{wind.loc}) and temporal (\texttt{wind}) dimension;Bureau of
Meteorology (BoM) provides climate observations that are widely applied
in agriculture and ecology study; air pollution data. \newline

\textbf{Our new data structure for spatio-temporal data}\\
This paper describes the implementation of a new spatio-temporal data
structure: \texttt{cubble}. \texttt{cubble} implements a relational data
structure that uses two forms to manage the switch between spatial and
temporal dimension. With this structure, users can manipulate the
spatial or temporal dimension separately, while leaves the linking of
two dimensions to \texttt{cubble}. The software is available from the
Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}. \newline

@@

The rest of the paper will be divided as follows: {[}complete when the
paper structure is more solid{]}
